### 2020-05-31 main.Rmd
Figures are now rendered from data (#1, #2, #3, #4)

```{r setup, include = FALSE}
library("papaja")
library(magrittr)
```

@@

```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf",
)
```

```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
)
```

@@

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
knitr::include_graphics("figures/model_arch.png")
```

```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
knitr::include_graphics("figures/final_train_progress.png")
```

```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")

```{r, confusionfinalmodels, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}

knitr::include_graphics("figures/final_confusion.png")

```{r, modelcorrs, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
knitr::include_graphics("figures/drafts.png")
```

```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")

```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

```{r, vsotherscfmat, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
knitr::include_graphics("figures/cf_other_tools.png")
```

```{r, activations, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
knitr::include_graphics("figures/convolution_maps.png")
```

@@

```{r, bacefftable}
apa_table(
  readr::read_tsv("tables/bacterial_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
)
```

```{r, fungiefftable}
apa_table(
  readr::read_tsv("tables/fungal_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
)
```

```{r, oomefftable}
apa_table(
  readr::read_tsv("tables/oomycete_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
)
```

```{r, possiblecombination}
apa_table(
  readr::read_tsv("tables/possible_combination_each_model.tab"),
  digits = 0,
  align = 'c',
  caption = 'List of possible combination of hyperparamaters setting for each model'
)
```

```{r, fixedhyperparam}
apa_table(
  readr::read_tsv("tables/fixed_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
)
```

```{r, tunedhyperparam}
apa_table(
  readr::read_tsv("tables/tuned_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
)
```

```{r, hyperparam }
 readr::read_tsv("tables/hyperparamscan_info.tab") %>% 
  knitr::kable( "latex", longtable = T, booktabs = T,
           digits = 0,
  align = 'c',
  caption = 'Hyperparameters used in automatic scanning of hyperparameter space and best performing parameters values for each model'
         ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)

>>>>

```{r setup, include = FALSE}
library("papaja")
library(dplyr)
library(ggplot2)
library(patchwork)

# Source plotting functions
source("functions/plot_confusion_mat.R")
source("functions/plot_confusion_matrices.R")
source("functions/get_lower_corr_mat.R")
source("functions/get_seqlogo.R")

# ggplot2 global theme
theme_set(theme_bw())
```

@@

```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf",
)
```

```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
)
```

@@

```{r effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
knitr::include_graphics("figures/model_arch.png")
```

```{r trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
knitr::include_graphics("figures/final_train_progress.png")
```

```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
```

```{r confusionfinalmodels, eval=FALSE, include=FALSE, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}

knitr::include_graphics("figures/final_confusion.png")
```

```{r modelcorrs, eval=FALSE, include=FALSE, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
knitr::include_graphics("figures/drafts.png")
```

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
# Load data
bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")

bacteria_test_label <- data.table::fread("data/raw/bacteria_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

oomycete_ensemble_results <- data.table::fread("data/raw/oomycete_pred_ens.csv",drop = "V1")

oomycete_test_label <- data.table::fread("data/raw/oomycete_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

fungi_ensemble_results <- data.table::fread("data/raw/fungi_pred_ens.csv", drop = "V1")

fungi_test_label <- data.table::fread("data/raw/fungi_testing.csv") %>%
  dplyr::select(label) %>%
  mutate_each(list(as.factor))

# Correlation plots
corr_bacteria <- get_lower_corr_mat(bacteria_ensemble_results)
corr_fungi <- get_lower_corr_mat(fungi_ensemble_results)
corr_oomycete <- get_lower_corr_mat(oomycete_ensemble_results)

# Confusion matrices
conf_bacteria <- plot_confusion_matrices(
  data = bacteria_ensemble_results,
  true_label = bacteria_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

conf_fungi <- plot_confusion_matrices(
  data = fungi_ensemble_results,
  true_label = fungi_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

conf_oomycete <- plot_confusion_matrices(
  data = oomycete_ensemble_results,
  true_label = oomycete_test_label,
  model_list = c("cnn_lstm", "cnn_gru", "gru_emb", "lstm_emb"),#, "ensemble_weighted", "ensemble_voting")
  fill_upper_limit = NA
)

# Patchwork
corr_patch <- (
  corr_bacteria + conf_bacteria +
    plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  ) /
  (
    corr_fungi + conf_fungi +
      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  ) /
  (
    corr_oomycete + conf_oomycete +
      plot_layout(ncol = 2, widths = c(1,3), tag_level = "new")
  )

corr_patch +
  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
```

```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
```

```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
# knitr::include_graphics("figures/cf_other_tools.png")

bacteria_confusion_mat_df <- readr::read_csv("data/processed/bacteria_confusion_mat_df.csv")
fungi_confusion_mat_df <- readr::read_csv("data/processed/fungi_confusion_mat_df.csv")
oomycete_confusion_mat_df <- readr::read_csv("data/processed/oomycete_confusion_mat_df.csv")

bacteria_conf <- plot_confusion_mat(
  bacteria_confusion_mat_df, 
  model_list = c("ensemble_weighted", "deepT3", "effectiveT3"), 
  fill_upper_limit = 40
)

fungi_conf <- plot_confusion_mat(
  fungi_confusion_mat_df, 
  model_list = c("cnn_lstm", "effectorp_1", "effectorp_2"),  
  fill_upper_limit = 40
)

oomycete_conf <- plot_confusion_mat(
  oomycete_confusion_mat_df,
  model_list = c("cnn_lstm", "effectR"),  
  fill_upper_limit = 40
)

(bacteria_conf + ggplot2::theme(legend.position = "none") +
    fungi_conf + 
    oomycete_conf + ggplot2::theme(legend.position = "none")) +
   plot_layout(ncol = 1, heights = c(1,1,1)) +
   plot_annotation(tag_levels = "A")
```

```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
# knitr::include_graphics("figures/convolution_maps_fixed.pdf")

all_bacteria <- rbind(
  data.table::fread("data/raw/bacteria_training.csv"),
  data.table::fread("data/raw/bacteria_validation.csv"),
  data.table::fread("data/raw/bacteria_testing.csv")
)

all_oomycete <- rbind(
  data.table::fread("data/raw/oomycete_training.csv"),
  data.table::fread("data/raw/oomycete_validation.csv"),
  data.table::fread("data/raw/oomycete_testing.csv")
)

all_fungi <- rbind(
  data.table::fread("data/raw/fungi_training.csv"),
  data.table::fread("data/raw/fungi_validation.csv"),
  data.table::fread("data/raw/fungi_testing.csv")
)

bacteria_heatmap <- data.table::fread("data/raw/bacteria_heatmap.csv")
fungi_heatmap <- data.table::fread("data/raw/fungi_heatmap.csv")
oomycete_heatmap <- data.table::fread("data/raw/oomycete_heatmap.csv")

bacteria_eff_seqlogo <- get_seqlogo(
  all_bacteria %>% dplyr::filter(label == 1), 
  method_plot = "bits", 
  start_pos = 25, 
  end_pos = 50, 
  break_size = 5
)

bacteria_non_eff_seqlogo <- get_seqlogo(
  all_bacteria %>% dplyr::filter(label == 0), 
  method_plot = "bits", 
  start_pos = 25, 
  end_pos = 50, 
  break_size = 5
)

oomycete_eff_seqlogo <- get_seqlogo(
  all_oomycete %>% dplyr::filter(label == 1), 
  method_plot = "bits", 
  start_pos = 15, 
  end_pos = 40, 
  break_size = 5
)

oomycete_non_eff_seqlogo <- get_seqlogo(
  all_oomycete %>% dplyr::filter(label == 0), 
  method_plot = "bits", 
  start_pos = 15, 
  end_pos = 40, 
  break_size = 5
)

fungi_eff_seqlogo <- get_seqlogo(
  all_fungi %>% dplyr::filter(label == 1), 
  method_plot = "bits", 
  start_pos = 35, 
  end_pos = 80, 
  break_size = 5
)

fungi_non_eff_seqlogo <- get_seqlogo(
  all_fungi %>% dplyr::filter(label == 0), 
  method_plot = "bits", 
  start_pos = 35, 
  end_pos = 80, 
  break_size = 5
)

heatmap_linebar_bacteria <- bacteria_heatmap %>% 
  dplyr::mutate(index = V1 + 1) %>% 
  dplyr::filter(index <= 750) %>% 
  dplyr::select(index, heatmap_bacteria_smooth) %>% 
  ggplot(aes(x = index, y = heatmap_bacteria_smooth)) +
  geom_line() +
  geom_vline(xintercept = 25, linetype = 'dashed', color = "red") +
  geom_vline(xintercept = 50, linetype = 'dashed',  color = "red") +
  scale_x_continuous(breaks = seq(0, 750, 150)) +
  labs(x = 'Position', y = "Value")

heatmap_linebar_oomycete <- oomycete_heatmap %>% 
  dplyr::mutate(index = V1 + 1) %>% 
  dplyr::filter(index <= 750) %>% 
  dplyr::select(index, heatmap_oomycete_smooth) %>% 
  ggplot(aes(x = index, y = heatmap_oomycete_smooth)) +
  geom_line() +
  geom_vline(xintercept = 15, linetype = 'dashed', color = "blue") +
  geom_vline(xintercept = 40, linetype = 'dashed', color = "blue") +
  geom_vline(xintercept = 45, linetype = 'dashed', color = "red") +
  geom_vline(xintercept = 70, linetype = 'dashed', color = "red") +
  scale_x_continuous(breaks = seq(0, 750, 150)) +
  labs(x = 'Position', y = "Value")

heatmap_linebar_fungi <- fungi_heatmap %>%
  dplyr::mutate(index = V1 + 1) %>% 
  dplyr::filter(index <= 750) %>% 
  dplyr::select(index, heatmap_fungi_smooth) %>% 
  ggplot(aes(x = index, y = heatmap_fungi_smooth)) +
  geom_line() +
  geom_vline(xintercept = 35, linetype = 'dashed', color = "blue") +
  geom_vline(xintercept = 80, linetype = 'dashed', color = "blue") +
  scale_x_continuous(breaks = seq(0, 750, 150)) +
  labs(x = 'Position', y = "Value")

seqlogo_patch <- (
  heatmap_linebar_bacteria + 
    bacteria_eff_seqlogo + theme(legend.position = "none") +
    bacteria_non_eff_seqlogo + theme(legend.position = "none") +
    plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
  ) /
  (
  heatmap_linebar_fungi + 
    fungi_eff_seqlogo + theme(legend.position = "none") +
    fungi_non_eff_seqlogo +
      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
  ) /
  (
  heatmap_linebar_oomycete + 
    oomycete_eff_seqlogo + theme(legend.position = "none") +
    oomycete_non_eff_seqlogo + theme(legend.position = "none") +
      plot_layout(ncol = 3, widths = c(2,3,3), tag_level = "new")
  )

seqlogo_patch +
  plot_annotation(tag_levels = c("A", "1"), tag_sep = ".")
```

@@

```{r bacefftable}
apa_table(
  readr::read_tsv("tables/bacterial_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
)
```

```{r fungiefftable}
apa_table(
  readr::read_tsv("tables/fungal_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
)
```

```{r oomefftable}
apa_table(
  readr::read_tsv("tables/oomycete_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
)
```

```{r possiblecombination}
apa_table(
  readr::read_tsv("tables/possible_combination_each_model.tab"),
  digits = 0,
  align = 'c',
  caption = 'List of possible combination of hyperparamaters setting for each model'
)
```

```{r fixedhyperparam}
apa_table(
  readr::read_tsv("tables/fixed_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
)
```

```{r tunedhyperparam}
apa_table(
  readr::read_tsv("tables/tuned_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
)
```

```{r hyperparam}
 readr::read_tsv("tables/hyperparamscan_info.tab") %>% 
  knitr::kable( "latex", longtable = T, booktabs = T,
           digits = 0,
  align = 'c',
  caption = 'Hyperparameters used in automatic scanning of hyperparameter space and best performing parameters values for each model'
         ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
